```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=4, fig.height=3, warning=FALSE, message=FALSE)
```

# Statistical Inference Course Project - Part 1
Author: Ashwin Venkatesh Prabhu

### 1. Overview:

In this project I will investigate the exponential distribution in R and compare it with the Central Limit Theorem. The exponential distribution can be simulated in R with rexp(n, lambda) where lambda is the rate parameter. The mean of exponential distribution is 1/lambda and the standard deviation is also 1/lambda. I will set lambda = 0.2 for all of the simulations. I will investigate the distribution of averages of 40 exponentials. Note that I will need to do a thousand simulations.

### 2. Simulations:

```{r}
library(ggplot2)

# set constants
lambda <- 0.2 # lambda for rexp
n <- 40 # # exponentials
noOfSimulations <- 1000 # number of tests

# set the seed to create reproducability
set.seed(23888)

expDistributions <- matrix(data=rexp(n * noOfSimulations, lambda), nrow=noOfSimulations)
expDistributionMean <- data.frame(means=apply(expDistributions, 1, mean))
```

```{r echo=FALSE}
# plot the means
ggplot(data = expDistributionMean, aes(x = means)) + 
  geom_histogram(binwidth=0.1) +   
  scale_x_continuous(breaks=round(seq(min(expDistributionMean$means), max(expDistributionMean$means), by=1)))
```

### 3. Sample Mean versus Theoretical Mean:

The expected mean of an exponential distribution of rate is: $\mu= \frac{1}{\lambda}$

```{r}
mu <- 1/lambda
mu
```

Let $\bar X$ be the average sample mean of 1000 simulations of 40 randomly sampled exponential distributions.

```{r}
meanOfMeans <- mean(expDistributionMean$means)
meanOfMeans
```
As you can see the expected mean and the avarage sample mean are very close

### 4. Sample Variance versus Theoretical Variance:

The expected standard deviation $\sigma$ of a exponential distribution of rate $\lambda$ is,
```{r}
sd <- 1/lambda/sqrt(n)
sd
```

The variance $Var$ of standard deviation $\sigma$ is
```{r}
variance <- sd^2
variance
```

Let $Var_x$ be the variance of the average sample mean of 1000 simulations of 40 randomly sampled exponential distribution, and $\sigma_x$ the corresponding standard deviation
```{r}
sd_x <- sd(expDistributionMean$means)
sd_x

variance_x <- var(expDistributionMean$means)
variance_x
```

As you can see the standard deviations are very close Since variance is the square of the standard deviations, minor differnces will we enhanced, but are still pretty close.

### 5. Distribution:

Comparing the population means & standard deviation with a normal distribution of the expected values. Added lines for the calculated and expected means

```{r}
# plot the means
ggplot(data = expDistributionMean, aes(x = means)) + 
  geom_histogram(binwidth=0.1, aes(y=..density..), alpha=0.2) + 
  stat_function(fun = dnorm, arg = list(mean = mu , sd = sd), colour = "red", size=1) + 
  geom_vline(xintercept = mu, size=1, colour="green") + 
  geom_density(colour="blue", size=1) +
  geom_vline(xintercept = meanOfMeans, size=1, colour="red") + 
  scale_x_continuous(breaks=seq(mu-3,mu+3,1), limits=c(mu-3,mu+3)) 
```

As you can see from the graph, the calculated distribution of means of random sampled exponantial distributions, overlaps quite nice with the normal distribution with the expected values based on the given lamba