```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, fig.width=8, fig.height=4)
options(width=100)
```

# Motor Trend - The relationship between a set of variables and miles per gallon
By Ashwin Venkatesh Prabhu

## Executive Summary

We work for Motor Trend, a magazine about the automobile industry. Looking at a data set of a collection of cars, we are interested in exploring the relationship between a set of variables and miles per gallon (MPG) (outcome).

- “Is an automatic or manual transmission better for MPG”
- "Quantify the MPG difference between automatic and manual transmissions"

We will start with the following steps:

- Process the data
- Conduct exploratory data analysis, focusing on the two paramaters we are interested in (Transmission and MPG)
- Model selection, where we try different models to help us answer our questions
- Model examination, to see wether our best model holds up to our standards
- A Conclusion where we answer the questions based on the data

## Processing

Change 'am' to a factor (1 = manual, 0 = automatic). Make cylinders a factor.
```{r}
library(ggplot2)
library(GGally)
library(dplyr)
library(ggfortify)

data(mtcars)

df <- mtcars
df$am <- as.factor(df$am)
levels(df$am) <- c("automatic", "manual")

df$cyl <- as.factor(df$cyl)
df$gear <- as.factor(df$gear)
df$vs <- as.factor(df$vs)
levels(df$vs) <- c("V", "S")
```

## Exploratory data analysis

Look at the dimensions & head of the dataset to get an idea
```{r}
# Result 1
dim(df)

# Result 2
head(df)
```

Let's take a look at the realtionship between the two parameters which we are intereseted in.
```{r}
plot1 <- ggplot(df, aes(am, mpg))
plot1 + geom_boxplot(aes(fill = am))
```

The above plot shows clearly that the manual transmissions have higher mpg's there could be a bias in the dataset that we are overlooking. 
Before creating a model we should look at which paramters to include besides 'am'. 
So we look at all correlations of parameters and take only those higher then the 'am' correlation.

```{r}
# Result 3
cors <- cor(mtcars$mpg, mtcars)
orderedCors <- cors[,order(-abs(cors[1,]))]
orderedCors

# Result 4
amPos <- which(names(orderedCors)=="am")
subsetColumns <- names(orderedCors)[1:amPos]
subsetColumns

df[,subsetColumns] %>% 
   ggpairs(
     mapping = ggplot2::aes(color = am), 
     upper = list(continuous = wrap("cor", size = 3)),
     lower = list(continuous = wrap("smooth", alpha=0.4, size=1), combo = wrap("dot"))
   )
```

## Model selection 

We have seen that mpg has many other (stronger) correlations than just 'am' we can guess that a model predicting the mpg solely on this parameter will not be the most accurate model. Let's check this out.

First we start with the basic model
```{r}
# Result 5
fit1 <- lm(mpg ~ am, df)
summary(fit1)
```

The p-values are actually quite low, the R-squared is problematic however. Now go to the other side of the spectrum by fitting all parameters of mtcars.

```{r}
# Result 6
fit2 <- lm(mpg ~ ., df)
summary(fit2)
```

The R-squared has improved, but the p-values hardly show any significance anymore. Perhaps this is due to overfitting. We now have to meet somewhere in the middle. Let's iterate using the step method.

```{r}
# Result 7
fit <- step(fit2, direction="both",trace=FALSE)
summary(fit)
```

## Model examination

The resulting best model `mpg ~ wt + qsec + am` is actually dependant on the transmission (am), but also weight (wt) and 1/4 mile time (qsec). All have significant p-values. The R-squared is pretty good to (0.85)

Now let's look (amongst others) at the Residuals vs Fitted 

```{r}
autoplot(fit)
```

The 'Normal Q-Q' plot looks ok, but the 'Residuals vs Fitted' and 'Scale-Location' both show worrysome trends.

## Conclusion

The question “Is an automatic or manual transmission better for MPG” can be answered because all models (#Result 5, #Result 6 and #Result 7) show that, holding all other paramters constant, manual transmission will increase your MPG.

The question “Quantify the MPG difference between automatic and manual transmissions” is harder to answer.

Based on the 'fit' (#Result 7) model `mpg ~ wt + qsec + am` we could conclude that (with a p < 0.05 confidence) cars with manual transmission have 2.9358 (say 3) more miles per gallon than automatic transmissions.
The model seems clean with a p < 0.05 and R squared of 0.85

The residuals vs fitted chart however warns us that there is something missing in our model. The real problem I think is that we only have 32 observations to train on (#Res1) and that observations hardly have overlap on the parameters 'wt' and 'qsec' (amongst others) if we look at the diagonal in the matrix chart

Although the conclusion of ca. 3 mpg better performance on manual transmissions seems feasible, I cannot with confidence conclude that this model will fit all future observations.